\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-09-22 analysis/paper/paper.tex
final version

\Abstract{
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician and the patient is the risk of an event given a
covariate profile. However, methods relying on time matching or risk-set
sampling (including Cox regression) eliminate the baseline hazard from
the likelihood expression or the estimating function and therefore needs
to be estimated separately using a non-parametric approach. This leads
to step-wise estimates of the cumulative incidence that are difficult to
interpret. Using case-base sampling, Hanley \& Miettinen (2009)
explained how the parametric hazard functions can be estimated using
logistic regression. Their approach naturally leads to estimates of the
cumulative incidence that are smooth-in-time.\\
In this paper, we present the \textbf{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can also be used in more complex
settings: competing risks, time-varying exposure, and variable
selection. Our package also includes an extensive array of visualization
tools to complement the analysis of time-to-event data. We illustrate
all these features through four different case studies.\\
*SRB and MT contributed equally to this work.
}

@@

Using the framework of logistic regression for analyzing time-to-event
data opens the door to an extensive array of powerful modeling tools.
Indeed, lasso and elastic-net regression can be used to select variables
that are associated with the hazard function. Logistic regression can
also be replaced by multinomial regression to analyse data with multiple
competing events.

@@

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \emph{et al} \citeyearpar{aalen2008survival}, we use the
placeholder \(\mathrm{past}\) to denote the past history of the
corresponding process. The reader interested in more details can refer
to Saarela \& Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
non-informative censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\]

@@

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest sampling the base series \emph{uniformly} from the study base.
In terms of Poisson processes, their sampling strategy corresponds
essentially to a time-homogeneous Poisson process with hazard equal to
\(\rho_i(t) = b/B\), where \(b\) is the number of sampled observations
in the base series, and \(B\) is the total population-time for the study
base (e.g.~the sum of all individual follow-up times). More complex
examples are also possible; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the intensity functions for the
sampling mechanism are proportional to the cardiovascular disease event
rate given by the Framingham score. With this sampling mechanism,
Saarela \& Arjas are able to increase the efficiency of their
estimators, when compared to uniform sampling.

@@

@@

To perform variable selection on the regression parameters
\(\theta \in \mathbb{R}^p\) of the hazard function, we can add a penalty
to the likelihood and solve the following problem: \begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
\end{equation} where \(\ell\left(\theta\right)\) is the log of the
likelihood function given in \eqref{eq:lik-function},
\(P(\theta_j;\lambda,\alpha)\) is a penalty term controlled by the
non-negative regularization parameters \(\lambda\) and \(\alpha\), and
\(w_j\) is the penalty factor for the \(j\)th covariate. These penalty
factors serve as a way of allowing parameters to be penalized
differently. For example, we could set the penalty factor for time to be
0 to ensure it is always included in the selected model.

@@

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our review to packages that implement at least
one of the following features: parametric modeling, non-proportional
hazard models, competing risk analysis, penalized estimation, and CIF
estimation. By searching for appropriate keywords in the
\code{DESCRIPTION} file of these packages, we found 60 relevant
packages. These 60 packages were then manually examined to determine
which ones are comparable to \pkg{casebase}. In particular, we excluded
packages that were focused on a different set of problems, such as
frailty and multi-state models. The remaining 14 packages appear in
Table \ref{tab:surv-pkgs}, along with some of the functionalities they
offer.

@@

\caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments). The points shown in the figure are select person-moments from the case and base series. We see that more events are observed at later follow-up times, motivating the use of a non-constant in time hazard model.}\label{fig:plot-stratified-erspc-data}
\end{figure}
\end{CodeChunk}

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by
first adding a layer for the area representing the population time in
gray, with subjects having the least amount of observation time plotted
at the top of the y-axis. We immediately notice a distinctive
\emph{stepwise shape} in the population time area. This is due to the
randomization of the Finnish cohorts which were carried out on January 1
of each of year from 1996 to 1999. Coupled with the uniform December 31
2006 censoring date, this led to large numbers of men with exactly 11,
10, 9 or 8 years of follow-up. Tracked backwards in time (i.e.~from
right to left), the population-time plot shows the recruitment pattern
from its beginning in 1991, and the January 1 entries in successive
years. Tracked forwards in time (i.e.~from left to right), the plot for
the first three years shows attrition due entirely to death (mainly from
other causes). Since the Swedish and Belgian centres were the last to
complete recruitment--in December 2003--the minimum potential follow-up
is three years. Tracked further forwards in time (i.e.~after year 3) the
attrition is a combination of deaths and staggered entries. As we can
see, population-time plots summarise a wealth of information about the
study into a simple graph.

@@

\[ CI_j(t) = \int_0^t f_j(u)du.\] The integrals are computed using
either numerical integration (via the trapezoidal rule) or Monte Carlo
integration. This option is controlled by the argument \code{method} of
the \code{absoluteRisk} function.

@@

\subsection{Different functional forms of time}

@@

\subsection{Time-dependent hazard ratios}

@@

\begin{CodeInput}
R> fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, data = ERSPC)
\end{CodeInput}
\end{CodeChunk}

In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from
prostate cancer hazard ratio and 95\% confidence interval for screening
vs.~control group as a function of time using the \code{plot} method for
objects of class \code{singleEventCB}:

@@

\caption[Estimated death from prostate cancer hazard ratio and 95\% confidence interval for screening vs]{Estimated death from prostate cancer hazard ratio and 95\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm. 95\% confidence intervals are calculated using the delta method. The plot shows that the cures attributable to the screening only begin to become statistically apparent by year 7 and later. The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
\end{figure}
\end{CodeChunk}

@@

In the next case study, we show how case-base sampling can be used in
the context of a competing risk analysis. For illustrative purposes, we
will use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is now available as part of the
\pkg{casebase} package.

@@

First, we can look at a population-time plot to visualize the incidence
density of both relapse and the competing events. In Figure
\ref{fig:compPop}, failure times are highlighted on the plot using red
dots for the event of interest (panel A) and blue dots for competing
events (panel B). In both panels, we see evidence of a non-constant
hazard function: the density of points is larger at the beginning of
follow-up than at the end.

@@

We will compare our hazard ratio estimates to that obtained from a Cox
regression. To do so, we need to treat the competing event as censoring.

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
+   data = bmtcrr
+ )
\end{CodeInput}
\end{CodeChunk}

@@

Given the estimate of the hazard function obtained using case-base
sampling, we can compute the absolute risk curve for a fixed covariate
profile. We perform this computation for a 35 year old woman who
received a stem-cell transplant from peripheral blood at relapse. We
compared the absolute risk curve for such a woman with ALL with that for
a similar woman with AML. We will estimate the curve from 0 to 60
months.

We will compare our estimates to that obtained from a corresponding
Fine-Gray model \citeyearpar{fine1999proportional}. The Fine-Gray model
is a semiparametric model for the cause-specific \emph{subdistribution
hazard}, i.e.~the function \(f_k(t)\) such that
\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
\(CI_k(t)\) is the cause-specific cumulative incidence. The Fine-Gray
model allows to directly assess the effect of a covariate on the
subdistribution hazard, as opposed to the hazard. For the computation,
we will use the \pkg{timereg} package \citep{timereg}:

@@

To illustrate this functionality, we use the dataset from the Study to
Understand Prognoses Preferences Outcomes and Risks of Treatment
(SUPPORT) \citep{knaus1995support}. The SUPPORT dataset tracks death in
five American hospitals within individuals who are considered seriously
ill.~The original data is available online from the Department of
Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
and imputed data consists of 9104 observations and 30 variables, and it
is available as part of the \pkg{casebase} package. In the comparisons
below, all covariates except \code{sps} and \code{aps} were modeled.
These two variables correspond to scores for predicting the outcome that
were developed as part of the original study. For more information about
this dataset, the reader is encouraged to look at the documentation in
our package.

For our penalized case-base model, we opted for the natural log of time
which corresponds to a Weibull distribution. For fitting the penalized
hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
to the \code{fitSmoothHazard} function. We must supply both a matrix
\code{y} containing the time and event variables, and a matrix \code{x}
containing all other covariates. We applied the lasso penalty by setting
\code{alpha = 1} and assigned a \code{penalty.factor} (\(w_j\)) of 0 to
the time variable to ensure it was in the selected model. We compared
our approach to both Cox regression, and lasso penalized Cox regression
(fitted via the \pkg{glmnet} package).

To compare the performance of our models, we split the data into 95\%
training and 5\% test sets. To assess both discrimination and
calibration, we used a time-dependent version of the classical Brier
score that is adjusted for censoring \citep{graf1999ass}. The Brier
score can be used with both parametric and semi-parametric models. We
used the \pkg{riskRegression} package to compute these scores for all
models.

@@

In Figure \ref{fig:cs3lolliPlot}, we show the coefficient estimates for
covariates that we selected by both penalized Cox and penalized
case-base. We note that both penalized approaches produce similar
results, and can clearly see the shrinkage effect owing to the
\(\ell_1\) penalty.

@@

We then compared the cumulative incidence estimation over the test set.
The probabilities over time for each observation are averaged, resulting
in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A).
We can see some minimal differences between the three models, with the
Kaplan-Meier giving the lowest estimates across follow-up-time. Note
that the apparent smoothness of the Cox and penalized Cox curves is due
to the large number of observations in the training set, which is used
to derive the Breslow estimate of the baseline hazard. As described
above, we compare the performance between the models by computing the
Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can
see that the Brier score is larger for the Kaplan-Meier estimate than
for the other three models. On the other hand, the differences between
these three models are minimal.

@@

Since the exposure is time-dependent, we need to manually define the
exposure variable \emph{after} case-base sampling and \emph{before}
fitting the hazard function. For this reason, we will use the
\texttt{sampleCaseBase} function directly.

@@

Next, we will compute the number of days from acceptance into the
program to transplant, and we use this variable to determine whether
each population-moment is exposed or not.

@@

\hypertarget{environment-details}{%
\section*{Environment Details}\label{environment-details}}
\addcontentsline{toc}{section}{Environment Details}

This report was generated on 2020-09-21 14:06:21 using the following
computational environment and dependencies:

\begin{CodeOutput}
#> R version 4.0.2 (2020-06-22)
#> Platform: x86_64-pc-linux-gnu (64-bit)
#> Running under: Pop!_OS 20.04 LTS
#> 
#> Matrix products: default
#> BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3
#> LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/liblapack.so.3
#> 
#> locale:
#>  [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
#>  [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
#>  [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
#>  [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
#>  [9] LC_ADDRESS=C               LC_TELEPHONE=C            
#> [11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] scales_1.1.1              forcats_0.5.0            
#>  [3] stringr_1.4.0             dplyr_1.0.2              
#>  [5] purrr_0.3.4               readr_1.3.1              
#>  [7] tidyr_1.1.2               ggplot2_3.3.2            
#>  [9] tidyverse_1.3.0           lubridate_1.7.9          
#> [11] visreg_2.7.0              tibble_3.0.3             
#> [13] riskRegression_2020.02.05 prodlim_2019.11.13       
#> [15] flexsurv_1.1.1            survival_3.2-3           
#> [17] pracma_2.2.9              magrittr_1.5             
#> [19] kableExtra_1.2.1          glmnet_4.0-2             
#> [21] Matrix_1.2-18             data.table_1.13.0        
#> [23] cowplot_1.1.0             colorspace_1.4-1         
#> [25] casebase_0.9.0           
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
#>  [4] htmlTable_2.1.0     base64enc_0.1-3     fs_1.5.0           
#>  [7] rstudioapi_0.11     MatrixModels_0.4-1  fansi_0.4.1        
#> [10] mvtnorm_1.1-1       xml2_1.3.2          codetools_0.2-16   
#> [13] knitr_1.29          Formula_1.2-3       jsonlite_1.7.1     
#> [16] broom_0.7.0         cluster_2.1.0       dbplyr_1.4.4       
#> [19] png_0.1-7           compiler_4.0.2      httr_1.4.2         
#> [22] backports_1.1.9     assertthat_0.2.1    cli_2.0.2          
#> [25] htmltools_0.5.0     quantreg_5.67       tools_4.0.2        
#> [28] gtable_0.3.0        glue_1.4.2          Rcpp_1.0.5         
#> [31] cellranger_1.1.0    vctrs_0.3.4         nlme_3.1-149       
#> [34] conquer_1.0.2       iterators_1.0.12    xfun_0.17          
#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.19   
#> [40] muhaz_1.2.6.1       MASS_7.3-53         zoo_1.8-8          
#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
#> [46] RColorBrewer_1.1-2  rticles_0.15        yaml_2.2.1         
#> [49] gridExtra_2.3       rms_6.0-1           rpart_4.1-15       
#> [52] latticeExtra_0.6-29 stringi_1.5.3       foreach_1.5.0      
#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.5        
#> [58] rlang_0.4.7         pkgconfig_2.0.3     matrixStats_0.56.0 
#> [61] evaluate_0.14       lattice_0.20-41     htmlwidgets_1.5.1  
#> [64] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
#> [67] R6_2.4.1            generics_0.0.2      Hmisc_4.4-1        
#> [70] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
#> [73] pillar_1.4.6        haven_2.3.1         foreign_0.8-79     
#> [76] mgcv_1.8-33         nnet_7.3-14         mstate_0.2.12      
#> [79] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
#> [82] timereg_1.9.7       jpeg_0.1-8.1        grid_4.0.2         
#> [85] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
#> [91] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
#> [94] quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}

The current Git commit details are:

\begin{CodeOutput}
#> Local:    editSahir /home/sahir/git_repositories/cbpaper
#> Remote:   editSahir @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
#> Head:     [dace5dd] 2020-09-20: added paragraph on variable selection in theoretical details
\end{CodeOutput}
\end{CodeChunk}

>>>>

\Abstract{
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician and the patient is the risk of an event given a
covariate profile. However, methods relying on time matching or risk-set
sampling (including Cox regression) eliminate the baseline hazard from
the likelihood expression or the estimating function. The baseline
hazard then needs to be estimated separately using a non-parametric
approach. This leads to step-wise estimates of the cumulative incidence
that are difficult to interpret. Using case-base sampling, Hanley \&
Miettinen (2009) explained how the parametric hazard functions can be
estimated using logistic regression. Their approach naturally leads to
estimates of the cumulative incidence that are smooth-in-time.\\
In this paper, we present the \textbf{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can also be used in more complex
settings: competing risks, time-varying exposure, and variable
selection. Our package also includes an extensive array of visualization
tools to complement the analysis of time-to-event data. We illustrate
all these features through four different case studies.\\
*SRB and MT contributed equally to this work.
}

@@

Logistic regression had already been used in the context of
discrete-time survival modeling \citep{cox1972regression}. But in the
context of continuous-time survival modeling, using the framework of
logistic regression opens the door to an extensive array of powerful
modeling tools. Indeed, lasso and elastic-net regression can be used to
select variables that are associated with the hazard function. Logistic
regression can also be replaced by multinomial regression to analyse
data with multiple competing events.

@@

\hypertarget{likelihood-and-estimating-function}{%
\subsection{Likelihood and estimating
function}\label{likelihood-and-estimating-function}}

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \emph{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
non-informative censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\]

@@

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest sampling the base series \emph{uniformly} from the study base.
In terms of Poisson processes, their sampling strategy corresponds
essentially to a time-homogeneous Poisson process with hazard equal to
\(\rho_i(t) = b/B\), where \(b\) is the number of sampled observations
in the base series, and \(B\) is the total population-time for the study
base (e.g.~the sum of all individual follow-up times). More complex
examples are also possible; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the intensity functions for the
sampling mechanism are proportional to the cardiovascular disease event
rate given by the Framingham score. Non-uniform sampling mechanisms can
increase the efficiency of the case-base estimators.

\hypertarget{common-parametric-models}{%
\subsection{Common parametric models}\label{common-parametric-models}}

@@

\hypertarget{competing-risk-analysis}{%
\subsection{Competing-risk analysis}\label{competing-risk-analysis}}

@@

\hypertarget{variable-selection}{%
\subsection{Variable selection}\label{variable-selection}}

To perform variable selection on the regression parameters
\(\theta \in \mathbb{R}^p\) of the hazard function, we can add a penalty
to the likelihood and optimise the following equation: \begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
\end{equation} where \(\ell\left(\theta\right) = \log L(\theta)\) is the
log of the likelihood function given in \eqref{eq:lik-function},
\(P(\theta_j;\lambda,\alpha)\) is a penalty term controlled by the
non-negative regularization parameters \(\lambda\) and \(\alpha\), and
\(w_j\) is the penalty factor for the \(j\)th covariate. These penalty
factors serve as a way of allowing parameters to be penalized
differently. For example, we could set the penalty factor for time to be
0 to ensure it is always included in the selected model.

@@

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citep{survTaskView}. For the purposes of this
article, we restricted our review to packages that implement at least
one of the following features: parametric modeling, non-proportional
hazard models, competing risk analysis, penalized estimation, and CIF
estimation. By searching for appropriate keywords in the
\code{DESCRIPTION} file of these packages, we found 60 relevant
packages. These 60 packages were then manually examined to determine
which ones are comparable to \pkg{casebase}. In particular, we excluded
packages that were focused on a different set of problems, such as
frailty and multi-state models. The remaining 14 packages appear in
Table \ref{tab:surv-pkgs}, along with some of the functionalities they
offer.

@@

\caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments). More events are observed at later follow-up times, motivating the use of non-constant hazard models.}\label{fig:plot-stratified-erspc-data}
\end{figure}
\end{CodeChunk}

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by
first adding a layer for the area representing the population time in
gray, with subjects having the least amount of observation time plotted
at the top of the y-axis. We immediately notice a distinctive
\emph{stepwise shape} in the population time area. This is due to the
randomization of the Finnish cohorts which were carried out on January 1
of each of year from 1996 to 1999. Coupled with the uniform December 31
2006 censoring date, this led to large numbers of men with exactly 11,
10, 9 or 8 years of follow-up. Tracked backwards in time (i.e.~from
right to left), the population-time plot shows the recruitment pattern
from its beginning in 1991, and the January 1 entries in successive
years. Tracked forwards in time (i.e.~from left to right), the plot for
the first three years shows attrition due entirely to death (mainly from
other causes). Since the Swedish and Belgian centres were the last to
complete recruitment in December 2003, the minimum potential follow-up
is three years. Tracked further forwards in time (i.e.~after year 3) the
attrition is a combination of deaths and staggered entries. As we can
see, population-time plots summarise a wealth of information about the
study into a simple graph.

@@

\[ CI_j(t) = \int_0^t f_j(u)du.\] Again, he integrals are computed using
either numerical integration (via the trapezoidal rule) or Monte Carlo
integration. This option is controlled by the argument \code{method} of
the \code{absoluteRisk} function.

@@

\hypertarget{different-functional-forms-of-time}{%
\subsection{Different functional forms of
time}\label{different-functional-forms-of-time}}

@@

\hypertarget{time-dependent-hazard-ratios}{%
\subsection{Time-dependent hazard
ratios}\label{time-dependent-hazard-ratios}}

@@

\begin{CodeInput}
R> fit_inter <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) * ScrArm, 
+                              data = ERSPC)
\end{CodeInput}
\end{CodeChunk}

In Figure \ref{fig:interaction-ERSPC}, we have the estimated hazard
ratio and 95\% confidence interval for screening vs.~control group as a
function of time using the \code{plot} method for objects of class
\code{singleEventCB}:

@@

\caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic B-spline basis of follow-up time and treatment arm. 95\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
\end{figure}
\end{CodeChunk}

@@

In the next case study, we show how case-base sampling can be used in
the context of a competing-risk analysis. For illustrative purposes, we
use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is now available as part of the
\pkg{casebase} package.

@@

First, we look at a population-time plot to visualize the incidence
density of both relapse and the competing events. In Figure
\ref{fig:compPop}, failure times are highlighted on the plot using red
dots for the event of interest (panel A) and blue dots for competing
events (panel B). In both panels, we see evidence of a non-constant
hazard function: the density of points is larger at the beginning of
follow-up than at the end.

@@

We want to compare our hazard ratio estimates to that obtained from a
Cox regression. To obtain these estimates, we need to treat the
competing event as censoring:

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
+                    data = bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

Given the estimate of the hazard function obtained using case-base
sampling, we can compute the absolute risk curve for a fixed covariate
profile. We perform this computation for a 35 year old woman who
received a stem-cell transplant from peripheral blood at relapse. We
compare the absolute risk curve for such a woman with ALL with that for
a similar woman with AML.

Next, we compare our estimates to that obtained from a corresponding
Fine-Gray model \citeyearpar{fine1999proportional}. The Fine-Gray model
is a semiparametric model for the cause-specific \emph{subdistribution
hazard}, i.e.~the function \(d_k(t)\) such that
\[CI_k(t) =1 - \exp\left( - \int_0^t d_k(u) \textrm{d}u \right),\] where
\(CI_k(t)\) is the cause-specific cumulative incidence. The Fine-Gray
model allows to directly assess the effect of a covariate on the
subdistribution hazard, as opposed to the cause-specific hazard. For the
computation, we use the \pkg{timereg} package \citep{timereg}:

@@

To illustrate this functionality, we use the dataset from the Study to
Understand Prognoses Preferences Outcomes and Risks of Treatment
(SUPPORT) \citep{knaus1995support}. The SUPPORT dataset tracks death in
five American hospitals within individuals who are considered seriously
ill.~The original data is available online from the Department of
Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
and imputed data consists of 9104 observations and 30 variables, and it
is available as part of the \pkg{casebase} package. In the comparisons
below, all covariates except \code{sps} and \code{aps} were used. These
two variables correspond to scores for predicting the outcome that were
developed as part of the original study. For more information about this
dataset, the reader is encouraged to look at the documentation in our
package.

For our penalized case-base model, we opt for the natural log of time
which corresponds to a Weibull distribution. For fitting the penalized
hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface
to the \code{fitSmoothHazard} function. We supply both a matrix \code{y}
containing the time and event variables, and a matrix \code{x}
containing all other covariates. We apply the lasso penalty by setting
\code{alpha = 1} and assign a \code{penalty.factor} (\(w_j\);
cf.~Equation \ref{eq:penest}) of 0 to the time variable to ensure it is
in the selected model. We compare our approach to both Cox regression,
and lasso penalized Cox regression (fitted via the \pkg{glmnet}
package).

To compare the performance of our models, we split the data into 95\%
training and 5\% test sets. To assess both discrimination and
calibration, we use a time-dependent version of the classical Brier
score that is adjusted for censoring \citep{graf1999ass}. The Brier
score can be used with both parametric and semi-parametric models. We
use the \pkg{riskRegression} package to compute these scores for all
models.

@@

In Figure \ref{fig:cs3lolliPlot}, we show the coefficient estimates for
covariates that we selected by both penalized Cox and penalized
case-base. We note that both penalized approaches produce similar
results. We can also clearly see the shrinkage effect owing to the
\(\ell_1\) penalty.

@@

We then compare the cumulative incidence estimation over the test set.
The probabilities over time for each observation are averaged, resulting
in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A).
We can see some minimal differences between the three models, with the
Kaplan-Meier giving the lowest estimates across follow-up-time. Note
that the apparent smoothness of the Cox and penalized Cox curves is due
to the large number of observations in the training set, which is used
to derive the Breslow estimate of the baseline hazard. As described
above, we compare the performance between the models by computing the
Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can
see that the Brier score is larger for the Kaplan-Meier estimate than
for the other three models. On the other hand, the differences between
these three models are minimal.

@@

Since the exposure is time-dependent, we need to manually define the
exposure variable \emph{after} case-base sampling and \emph{before}
fitting the hazard function. For this reason, we use the
\texttt{sampleCaseBase} function directly.

@@

Next, we compute the number of days from acceptance into the program to
transplant, and we use this variable to determine whether each
population-moment is exposed or not.

@@
